---
slug: "bookdown-learnings"
title: 10 bookdown Things we Learned by Working on the Blog Guidance
authors:
  - Stefanie Butland
  - Maëlle Salmon
# Set the date below to the publication date of your post
date: 2020-03-01
# Set categories to technotes if this is a tech note
categories: blog
# Leave topicid blank below; will be set by editor
topicid:
# Minimal tags for a post about a community-contributed package 
# that has passed software peer review are listed below
# Consult the Technical Guidelines for information on choosing tags
tags:
  - community
  - bookdown
# the summary below will be used by e.g. Twitter cards
description: "10 bookdown tips for newbies and experts"
# If you have no preferred image for Twitter cards,
# delete the twitterImg line below 
# Note there is no '/' symbol before 'img' here
# if needed replace blog with technotes
#twitterImg: blog/2019/06/04/post-template/name-of-image.png
# 'output' is necessary to obtain index.md
# Do not commit index.html
output: 
  html_document:
    keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path = "", comment = "", warning = FALSE)

knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)
```

Blablabla blog guidance blablabla. Technically, we structured the content as a bookdown gitbook which [Karthik Ram](/authors/karthik-ram/) judiciously suggested us. It was Stef's first foray into the glorious process of publishing a book with bookdown, and Maëlle's second one[^1].

### Stef (first-timer)


### Maëlle (more experienced bookdowner)

#### 1. How to start a bookdown project

I've started both my bookdown projects using [Sean Kross' excellent primer](https://github.com/seankross/bookdown-start),
but whilst looking for a reference to show Stef[^2],
I saw a [tweet of Alison Hill's about starting a bookdown project from RStudio](https://twitter.com/apreshill/status/1137183351125577728) which looks handy. 
[Slide of Alison's showing how to start a bookdown from RStudio](https://arm.rbind.io/slides/bookdown.html#11).

#### 2. How to get the copy-paste button for code chunks

In the blog guidance, if you hover around the top-right corner of [e.g. the Markdown post template](https://blogguide.ropensci.org/templatemd.html) you get a copy-paste button. 
For this to work, the chunk needs to have some language information i.e.

````
```
code
```
````

will not get a copy-paste button, but 

````
```yaml
code
```
````

will! 
I'm glad I know that now. 
Chunks with language info are prettier anyway since they get adapted code highlighting.

#### 3. How to define functions and chunk options for all chapters

I found how to write code that'll be loaded before each chapter rendering thanks to [Christophe Dervieux's answer on an old RStudio community thread](https://community.rstudio.com/t/bookdown-caching/43652/2).

One simply needs to create an [R script called `_common.R` at the root of the bookdown project](https://raw.githubusercontent.com/ropensci-org/blog-guidance/master/_common.R). 
Here's ours below. 
It contains chunk options, `magrittr` loading, and two helper functions for rendering templates.

```{r commonr, echo = FALSE}
details::details(readLines("https://raw.githubusercontent.com/ropensci-org/blog-guidance/master/_common.R"), summary = "our _common.R")
```

Our bookdown project uses `DESCRIPTION` to track dependencies, I suppose I could use the package infrastructure more and define the helper functions as functions _of a package_, but the approach above is pleasant too.

#### 4. How to have alternative text but no captions for figures

In R Markdown, the same chunk options `fig.cap` controls the caption and alternative text of images.
We wanted [alternative text](https://www.w3.org/WAI/tutorials/images/informative/) but no caption.
The header option `figure_caption` didn't work.
I opened [an issue in bookdown repo](https://github.com/rstudio/bookdown/issues/856) after not getting solutions [on an RStudio community post](https://community.rstudio.com/t/have-an-alternative-text-but-no-caption-or-a-different-caption-in-bookdown-git-book/50370).

So if bookdown doesn't really support having alternative text but no captions for figures, what did we do? Thanks to a good tip by [Romain Lesur](https://github.com/RLesur), I wrote CSS to remove all captions.

```css
.caption {
  display: none;
}
```

The lines above live with other styling stuff in a file called `style.css`, that we refer to in [the `_output.yml` config file](https://github.com/ropensci-org/blog-guidance/blob/cae5887f392b4137eec50467779e9bba973b0b69/_output.yml#L3).

#### 5. How to deploy a preview of the book for PRs

I felt quite strongly about having some sort of CI/CD for the book. We achieved that using GitHub Actions, a new CI service by GitHub. If you're curious about it, I'd recommend watching [Jim Hester's talk from the RStudio conference earlier this year](https://resources.rstudio.com/rstudio-conf-2020/azure-pipelines-and-github-actions-jim-hester). Our GitHub Actions workflows make good use of [Jim Hester's actions and examples](https://github.com/r-lib/actions).

Here's what we now have

* after each commit to the master branch, the book is built and deploy to the gh-pages branch that points to the blog guide official URL. Code below, [log example](https://github.com/ropensci-org/blog-guidance/runs/471978792?check_suite_focus=true);

```{r masterworkflow, echo = FALSE}
details::details(readLines("https://raw.githubusercontent.com/ropensci-org/blog-guidance/master/.github/workflows/master.yml"), summary = "master workflow", lang = "yaml")
```

* after each commit in a PR from a fork, the book is built so we'd notice something breaking the Rmd. [log example](https://github.com/ropensci-org/blog-guidance/runs/466995315?check_suite_focus=true);

* after each commit in a PR from the repo, the book is built and deployed to a Netlify preview whose URL is posted in the PR checks. [Log example](https://github.com/ropensci-org/blog-guidance/runs/471968808?check_suite_focus=true), [Direct link to the check with the preview link](https://github.com/ropensci-org/blog-guidance/runs/471974005?check_suite_focus=true).

```{r prworkflow, echo = FALSE}
details::details(readLines("https://raw.githubusercontent.com/ropensci-org/blog-guidance/master/.github/workflows/pr.yml"), summary = "PR workflow", lang = "yaml")
```

Highlights from the PR workflow above

* To deploy to Netlify _and_ get the preview URL, the workflow doesn't use Netlify's GitHub Actions but instead installs Netlify CLI, extracts the URL using `jq`[^3] and set it as an environment variable that can be used by next steps. I got the idea from [a thread on Netlify forum](https://community.netlify.com/t/deploying-preview-web-hook-via-api/3320/2).

```yaml
run: DEPLOY_URL=$(netlify deploy --site ${{ secrets.NETLIFY_SITE_ID }} --auth ${{ secrets.NETLIFY_AUTH_TOKEN }} --dir=docs --json | jq '.deploy_url' --raw-output);echo "::set-env name=DEPLOY_URL::$DEPLOY_URL"
```

* To post the preview URL in the PR, the workflow uses [a GitHub check run](https://developer.github.com/v3/checks/runs/). You can only create and patch them as a GitHub app. Luckily when in a GitHub Actions runner the default token has these permissions! My only regret is that trying to post the preview URL as a "details" URL for that check didn't work (GitHub resets it to the check page on GitHub Actions) but it's not too bad.

After the successful deployment of a preview, in the PR checks one check called "Preview book" appears.

<!--html_preserve--> {{< figure src = "PR-checks.png" width = "600" alt = "Screenshots of PR checks in a PR to the blog guide" >}}<!--/html_preserve-->

If clicking on "Details" one get to [a check page where the preview link is prominent](https://github.com/ropensci-org/blog-guidance/runs/471974005?check_suite_focus=true). 

<!--html_preserve--> {{< figure src = "preview-check.png" width = "600" alt = "Screenshots of a GitHub Action check page featuring a Netlify preview link" >}}<!--/html_preserve-->


* To find out whether the PR is from a fork (and to skip all deploy steps for that, since forks don't have access to the Netlify secrets saved in our repo settings), I use `jq` on the raw info about the build, idea I got [from Vanessa Sochat](https://github.com/r-universe/vsoch/blob/2bf10b1e30f59f5fab64ec2b7332526c47f1f4d3/.github/workflows/pull-request-update-packages.yml#L26).

```yaml
      - name: Is this a fork
        run: |
          fork=$(jq --raw-output .pull_request.head.repo.fork "${GITHUB_EVENT_PATH}");echo "::set-env name=fork::$fork"
```

And some of the further steps are skipped based on `fork`.

```yaml
      - name: Install Netlify CLI
        if: env.fork == 'false'
        run: npm install netlify-cli -g
```

If you're feeling motivated to add GitHub Actions deployment to your bookdown project, on top of Jim's video mentioned earlier and our workflows, be sure to note that Emil Hvitfeld wrote an [excellent guide to deploying your book on Netlify with GitHub Actions](https://www.hvitfeldt.me/blog/bookdown-netlify-github-actions/), with screenshots!

### Conclusion

In this post, we shared useful bookdown things we learnt whilst working on rOpenSci blog guidance. 
We're quite happy with both the end product and our learnings! 
If _you_ want to get started with bookdown, we'd recommend [the bookdown book](https://bookdown.org/yihui/bookdown/introduction.html), [this introductory slidedeck by Alison Hill](https://alison.rbind.io/talk/2019-rsc-bookdown/) and [the bookdown gallery](https://bookdown.org/home/archive/).


[^1]: The first bookdown book Maëlle worked on is [rOpenSci dev guide](https://devguide.ropensci.org/)!
[^2]: [A tweet by Hadley Wickham about using the config file for ordering chapters rather than numbering their filenames](https://twitter.com/hadleywickham/status/1137317951428747270)
[^3]: Maëlle discovered `jq` in [a blog post by Carl Boettiger](https://www.carlboettiger.info/2017/12/11/data-rectangling-with-jq/) and reported on her use of `jq` with rOpenSci `jqr` R package [in a blog post about getting data about Software Peer Review](/blog/2018/04/26/rectangling-onboarding/).
